---
title: "Plane Crashes Data Analysis"
author: "Nanalise Howe"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Load requires libraries

```{r}
library(readr)
library(tidyverse)
library(lubridate)
library(ggplot2)
```


# Loading Data
```{r}
crashes <- read_csv("aircraft_crash_data.csv")

# quick look at structure
glimpse(crashes)
summary(crashes)
```


# Data Cleaning and Preparation
# Convert accident date to Date format, extract Year, and convert fatalities to numeric
```{r}
crashes <- crashes %>%
  mutate(Date = as.Date(`acc. date`, format = "%d-%b-%y")) %>%
  mutate(Year = year(Date)) %>%
  mutate(fat. = na_if(fat., "")) %>%
  mutate(fat. = as.numeric(fat.)) %>%
  filter(!is.na(Year) & Year >= 1900 & Year <= 2024)
```


# Remove rows with missing key info columns
# Calculate total crashes and fataliteis, average fatalities per crash

```{r}
key_cols <- c("acc. date", "type", "reg.", "operator", "fat.", "location", "dmg")
crashes_clean <- crashes[complete.cases(crashes[, key_cols]), ]

summary(crashes_clean)
```

# Summary Statistics

```{r}
summary_stats <- crashes_clean %>%
  summarise(
    total_crashes = n(),
    total_fatalities = sum(fat., na.rm = TRUE),
    avg_fatalities_per_crash = mean(fat., na.rm = TRUE)
  )

summary_stats
```


# Create crashes per year

```{r}
crashes_per_year <- crashes_clean %>%
  group_by(Year) %>%
  summarise(
    crashes = n(),
    fatalities = sum(fat., na.rm = TRUE)
  )

```

# Plotcrashes and fatalities trend

```{r}

library(ggplot2)

ggplot(crashes_per_year, aes(x = Year)) +
  geom_line(aes(y = crashes), color = "blue", size = 1) +
  geom_line(aes(y = fatalities), color = "red", size = 1) +
  labs(title = "Crashes and Fatalities per Year",
       y = "Count",
       x = "Year",
       caption = "Blue = Crashes, Red = Fatalities") +
  theme_minimal()

```

# Crashes by Aircraft Type
```{r}
crashes_by_type <- crashes_clean %>%
  group_by(type) %>%
  summarise(
    crashes = n(),
    fatalities = sum(fat., na.rm = TRUE)
  ) %>%
  arrange(desc(crashes))

# Show top 10 aircraft types by number of crashes
head(crashes_by_type, 10)
```
# Top Operators by Fatalities
```{r}
# Summarize crashes and fatalities by operator
top_operators <- crashes_clean %>%
  group_by(operator) %>%
  summarise(
    crashes = n(),
    fatalities = sum(fat., na.rm = TRUE)
  ) %>%
  arrange(desc(fatalities)) %>%
  head(10)

top_operators

```

